Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations

### 2020-05-07 paper/tidy-missing-data-paper.Rmd
done proofing the graphics section

Missing values are often ignored when plotting data - so much so that the data visualisation software, MANET, is an acronym for "Missings Are Now Equally Treated" [@Unwin1996]. However, plots can help to reveal the type of missing value patterns, even those of MCAR, MAR or MNAR. Here we summarise how to systematically explore missing patterns visually, and define useful plots to make, relative to the nabular data structure.

@@

The shadow matrix allows both the variables and cases to be displayed using a heatmap style visualization, `vis_miss()` from \pkg{visdat} [@visdat]. This also provides numerical summaries of missingness in the legend, and for each column (Figure \@ref(fig:gg-miss-case-var)C-D). Clustering can be applied to the rows, and columns arranged in order of missingness (Figure \@ref(fig:gg-miss-case-var)D). Similar visualizations are available in other packages such as \pkg{VIM}, \pkg{mi}, \pkg{Amelia}, and \pkg{MissingDataGUI}. A key improvement is `vis_miss()` orients the visualization analogous to a regular data structure: variables form columns and are named at the top, and each row is an observation. Using \pkg{ggplot2} makes the plot easily customizable.

@@

The number of times observations are missing together is visualized using an "upset plot" [@Conway2017]. An alternative to a Venn diagram, an upset plot shows the size and features of overlapping sets, scaling well with more variables. An upset plot can be constructed from the shadow matrix, with Figure  \@ref(fig:airquality-upset) visualising overlapping amounts missings in the airquality data. The bottom right shows the combinations of missingness, the top panel shows the size of these combinations, and the bottom left shows missingness in each variable. This provides similar information to Figure \@ref(fig:gg-miss-case-var)A-D, but more clearly illustrating overlapping missingness, where 2 cases are missing together in variables Solar.R and Ozone.

@@

Missing values are by default not shown for univariate visualizations such as histograms or densities. Two ways to use _nabular_ data to present univariate data with missings are discussed. The first imputes values below the range to facilitate visualizations. The second displays two plots of the same variable according to the missingness of a chosen variable.

@@

**Imputing values below the range**. To visualize the amount of missings in each variable, the data is transformed into _nabular_ form, then values are imputed below the range of data using `impute_below`. Figure \@ref(fig:impute-shift-histogram)A shows a histogram of Ozone values on the right in green, and the  count of missing ozone values on the left, in orange. The missings in Ozone are mapped to the fill aesthetic in ggplot by referring to the variable `Ozone` as `Ozone_NA`, the values imputed below the range. Nabular data facilitates adding counts of missingness to a histogram, allowing not only understanding of a variables' distribution of values, but also understanding of the magnitude of missings.

**Univariate split by missingness**. Two distributions of a variable can be shown according to the missingness of another variable. Figure \@ref(fig:impute-shift-histogram) shows the values of temperature when ozone is present, and missing. Figure \@ref(fig:impute-shift-histogram)A is a faceted histogram, and Figure \@ref(fig:impute-shift-histogram)B is an overlaid density. This shows how values of temperature are affected by the missingness of ozone, and reveals a cluster of low temperature observations with missing ozone values.
The nabular structure of the data helps reveals the distribution of the missing values. This can facilitate exploring missing data distributions. For example, we would expect if data were MCAR, for values to be roughly uniformly missing thoughout the histogram or density.

@@

Parallel coordinate plots can visualize missingness beyond two dimensions. They transform variables to the same scale, ranging between 0 and 1. The `oceanbuoys` dataset from \pkg{naniar} is used for this visualisation, containing measurements of sea and air temperature, humidity, and east west and north south wind speeds. Data was collected in 1993 and 1997, to understand and predict El Niño and El Niña. Figure \@ref(fig:parallel-cord-plot) is a parallel coordinate plot of `oceanbuoys`, with missing values imputed to be 10% below the range, and values colored according to whether humidity was missing (`humidity_NA`). Figure \@ref(fig:parallel-cord-plot) shows humidity is missing at low air and sea temperatures, and humidity is missing in one year, and one location.

>>>>

Missing values are often ignored when plotting data - which is why the data visualisation software, MANET, was named using an acronym for "Missings Are Now Equally Treated" [@Unwin1996]. However, plots can help to reveal the type of missing value patterns, even those of MCAR, MAR or MNAR. Here we summarise how to systematically explore missing patterns visually, and define useful plots to make, relative to the nabular data structure.

@@

The shadow matrix allows both the variables and cases to be displayed using a heatmap style visualization, with `vis_miss()` from \pkg{visdat} [@visdat]. This also provides numerical summaries of missingness in the legend, and for each column (Figure \@ref(fig:gg-miss-case-var)C-D). Clustering can be applied to the rows, and columns arranged in order of missingness (Figure \@ref(fig:gg-miss-case-var)D). Similar visualizations are available in other packages such as \pkg{VIM}, \pkg{mi}, \pkg{Amelia}, and \pkg{MissingDataGUI}. A key improvement is `vis_miss()` orients the visualization analogous to a regular data structure: variables form columns and are named at the top, and each row is an observation. Using \pkg{ggplot2}, as the foundation, makes the plot easily customizable.

@@

The number of times observations are missing together can be visualized using an "upset plot" [@Conway2017]. An alternative to a Venn diagram, an upset plot shows the size and features of overlapping sets, and scales well with more variables. An upset plot can be constructed from the shadow matrix, as shown in Figure  \@ref(fig:airquality-upset) which displays the overlapping counts of  missings in the airquality data. The bottom right shows the combinations of missingness, the top panel shows the size of these combinations, and the bottom left shows missingness in each variable. This provides similar information to Figure \@ref(fig:gg-miss-case-var)A-D, but more clearly illustrating overlapping missingness, where 2 cases are missing together in variables Solar.R and Ozone.

@@

Missing values are typically not shown for univariate visualizations such as histograms or densities. Two ways to use _nabular_ data to present univariate data with missings are discussed. The first imputes values below the range to facilitate visualizations. The second displays two plots of the same variable according to the missingness of a chosen variable.

@@

**Imputing values below the range**. To visualize the amount of missings in each variable, the data is transformed into _nabular_ form, then values are imputed below the range of data using `impute_below_all`. Figure \@ref(fig:impute-shift-histogram)A shows a histogram of Ozone values on the right in green, and the histogram of missing ozone values on the left, in orange. The missings in Ozone are imputed and `Ozone_NA` is  mapped to the fill aesthetic. Nabular data facilitates adding counts of missingness to a histogram, allowing examination of a variables' distribution of values, and also the magnitude of missings.

**Univariate split by missingness**. The distribution of a variable can be shown according to the missingness of another variable. The shadow matrix part of nabular is used to handle the faceting, and color mapping. Figure \@ref(fig:impute-shift-histogram) shows the values of temperature when ozone is present, and missing, a faceted histogram (B), and  an overlaid density (C).  This shows how values of temperature are affected by the missingness of ozone, and reveals a cluster of low temperature observations with missing ozone values. This type of plot can facilitate exploring missing data distributions. For example, we would expect if data were MCAR, for values to be roughly uniformly missing thoughout the histogram or density.

@@

Parallel coordinate plots can help to visualize missingness beyond two dimensions. They transform variables to the same scale, ranging between 0 and 1. The `oceanbuoys` dataset from \pkg{naniar} is used for this visualisation, containing measurements of sea and air temperature, humidity, and east west and north south wind speeds. Data was collected in 1993 and 1997, to understand and predict El Niño and El Niña. Figure \@ref(fig:parallel-cord-plot) is a parallel coordinate plot of `oceanbuoys`, with missing values imputed to be 10% below the range, and values colored according to whether humidity was missing (`humidity_NA`). Figure \@ref(fig:parallel-cord-plot) shows humidity is missing at low air and sea temperatures, and humidity is missing in one year, and one location.
